Exploring the tree of numerical semigroups

### 2014-09-04 Papier/article.tex
Add goal of the paper in the introduction and do some stuff in last section

\documentclass[reqno,11pt]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{pgfplots}
\theoremstyle{plain}
\newtheorem{prop}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{defi}[prop]{Definition}
\newtheorem{nota}[prop]{Notation}
\newtheorem{exam}[prop]{Example}
\newtheorem{lem}[prop]{Lemma}
\newtheorem{fact}[prop]{Fact}

@@

\newcommand{\gr}[1]{{\color{gray} #1}}
%\newcommand{\card}{\#}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\kkp}{{k+1}}
\newcommand{\gen}[1]{\left<#1\right>}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\newcommand{\NN}{\mathbb{N}}
\renewcommand{\tt}[1]{\texttt{#1}}
\newcommand{\sgnode}[1]{{\bf \left<#1\right>}}
\renewcommand{\ie}{\emph{i.e}}
\DeclareMathOperator{\Irr}{Irr}
\DeclareMathOperator{\App}{App}
\DeclareMathOperator{\card}{card}

\newcommand{\TODO}[1]{\marginpar{\small To do: #1}}

@@

M.~Bras-Amor\'os used a depth first search exploration of the tree of numerical semigroups up to a given genus.
This tree was introduced by J.C.~Rosales and al. in \cite{Rosales} and it is the subject of the Section~\ref{S:Tree}.
Starting with all the numerical semigroups of genus $49$ she obtained the number of numerical semigroups of genus $50$ in $18$ days on a pentium D runing at $3$GHz. 
In the package~\tt{NumericalSgs} \cite{NumericalSgps} of \tt{GAP} \cite{GAP}, M. Delgado together with P.A.~Garcia-Sanchez and J. Morais used the same method of exploration.

\TODO{Present the goal of the paper.}

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
In Section~\ref{S:DecNumber} we describe a new representation of numerical semigroups that is well suited to the construction of the tree. 
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimisation of the algorithm introduced in Section~\ref{S:Algo}.

@@

To get the greatest speed from modern processor, we used several optimization
tricks, which we will elaborate in the following section:
\begin{itemize}
\item Vectorization (MMX, SSE instructions sets) and careful memory alignment.
\item Shared memory multi-core computing using \texttt{CILK++} for low level
  enumerating tree branching;
\item Partially derecursived algorithm using a stack;
\item Avoiding all dynamic allocation during the computation: everything is
  computed ``in place'';
\item Avoiding all unnecessary copy (the cost of the \textsc{Son} algorithm is
  roughly the same as copying) !
\item Aggressive loop unrolling: the main loop is unrolled by hand using some
  kind of Duff's device.
\item Carefull choice of data type (\verb|uint_fast8_t| for decomposition
  number, vs \verb|uint_fast64_t| for all indexes).
\end{itemize}

@@

Assume for example that we want to construct the tree $\mathcal{T}^{\leq 100}$
of all numerical semigroup of genus smaller than $100$.  In this case, the
representation of numerical semigroup given in Section~\ref{S:DecNumber} uses
decomposition numbers of integers smaller than~$300$.  By Lemma~\ref{L:RepN},
such a decomposition number is smaller than $151$ and requires~$1$ byte of
memory.  Thus at each \textbf{for} step of Algorithm~\textsc{Son}, the CPU
actually works on $1$ byte. However current CPUs usually work on $8$ bytes and
even on $16$~bytes using vector extensions. The first optimization uses this
point.

@@

\noindent In this version we can see that the cursors \texttt{src} and
\texttt{dst} move at the same time and that the modification of the value
pointed by \texttt{dst} only needs to access the values pointed by \tt{src}
and \tt{dst}.  We can therefore work in multiple entries at the same time
without collision.  Current CPUs allow this thanks to the SIMD technologies as
MMX, SSE, etc.  The acronym SIMD~\cite{SIMD} stands for Single Operation
Multiple Data.  We used SSE4.1~\cite{SSE, IntelSSE} technology as it allows
for the largest speedup. This need to respect some constraints in the memory
organization of the data, namely ``memory alignment''. Recall that an address
is 16~bytes aligned if it is a multiple of 16. SSE memory access are much
faster for aligned access.

The computation of the children is then performed as follows. First, the
parent's decomposition numbers are copied in the children's using the
following C++ code
\begin{listing}{1}
void copy_blocks(dec_blocks &dst, dec_blocks const &src) {
  for (ind_t i=0; i<NBLOCKS; i++) dst[i] = src[i];
}
\end{listing}
Here \tt{dec\_blocks} is a type for arrays of 16~bytes blocks whose size
\tt{NBLOCKS} are just large enough to store the decomposition numbers (that is
$3G$ rounded up to a multiple of 16). The instruction \tt{dst[i] = src[i]}
actually copy a full 16~bytes block.

Then the core of the \textbf{while} loop in the preceding algorithm is
translated as a for loop as follows (recall that $x$ denotes the generator of
the father which is to be removed in the children):
\begin{listing}{1}
start = x >> 4;      // index of the block containing x
shift = x & 0xF;     // offset of x inside the block
...                  // some specific instructions to
...                  // handle the first incomplete block.
for (long int i=start+1; i<NBLOCKS; i++) {
  block = load_unaligned_epi8(src + ((i-start)<<4) - shift);
  dst[i] -= ((block != zero) & one);
}
\end{listing}
The instruction \tt{load\_unaligned\_epi8} (specific to SSE technology) loads
in the variable \tt{block} 16 consecutive entries of the decomposition number
of the parent (called \tt{src} monoid). Those entries will be used to compute
the entries $16i, \dots 16i+15$ of the children monoid. Since the removed
generator $x$ is not necessarily a multiple of $16$, the data are not aligned
in memory, hence the use of a specific instruction. The \tt{zero}
(resp. \tt{one}) constants are initialized as 16 bytes equal to $0$
(resp. $1$). The comparison \tt{(block != zero)} therefore returns a block
which contains $0$ in the bytes corresponding the $0$ entries of block and
$255$ in the non zero one. This results is then bitwise anded with one so that
the instruction actually performs a 16~bytes parallel version of
\begin{algorithmic}
\State $\tt{dst} \gets \tt{dst} -
\textbf{ if } \tt{block} \neq 0 \textbf{ then } 1 \textbf{ else } 0$
\end{algorithmic}
which is equivalent to Lines \tt{10} to \tt{14} of Algorithm~\textsc{Son}.
\medskip

@@

\subsection{Parallel tree exploration using \texttt{Cilk+}}

Our second optimization is to use parallelism on exploration of the tree.
Today, CPU of personal computer have several cores (2, 4 or more).  The given
version of our exploration algorithm uses a single core and so a fraction only
of the power of a CPU. The idea here is that different branches of the tree can
be explored in parallel by different cores of the computer. The tricky part is
to ensure that each cores are busy, giving a new branch when a core is done
with a former one. Fortunately there is a technology called
\texttt{Cilk+}~\cite{CilkIntel, CilkRefman}
which is particularly well suited for those kind of problems.

\tt{Cilk} is a general-purpose language designed for multithreaded parallel
computing. The \tt{C++} incarnation is called \tt{Cilk+}. The biggest
principle behind the design of the Cilk language is that the programmer should
be responsible for \emph{exposing} the parallelism, identifying elements that
can safely be executed in parallel; it should then be left to the run-time
environment, particularly the scheduler, to decide during execution how to
actually divide the work between processors.

The crucial thing is that two keywords are all that are needed to start
using the parallel features of \tt{Cilk+}:
\begin{itemize}
\item\tt{cilk\_spawn}: used on a procedure call, indicates that the call can
  safely operate in parallel with the remaining code if the current
  function. Note that the scheduler is not obligated to run this procedure in
  parallel; the keyword merely alerts the scheduler that it can do so.
\item\tt{cilk\_sync}: indicates that execution of the current procedure cannot
  proceed until all previously spawned procedures have completed and returned
  their results to the current frame.
\end{itemize}
As a consequence, to get a parallel version of the recursive
Algorithm~\ref{A:ExpRec}, you only need to modify it as
\begin{algorithmic}[1]
\Procedure{ExploreRec}{\tt{S}, \tt{G}}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State\textbf{CilkSpawn} \Call{ExploreRec}{$\tt{S}^\tt{x}, \tt{G}$}
			\EndIf
		\EndFor
	\EndIf
\EndProcedure
\end{algorithmic}
We just tells \tt{Cilk+} that the subtrees rooted at various children can be
explored in parallel.

The thing are actually only a little bit more complicated. First we have to
gather the results of the exploration. If we simply write
\begin{algorithmic}[1]
\State \tt{result}[$g(\tt{S}$] $\gets \tt{result}[g(\tt{S}] + 1$
\end{algorithmic}
then we face the problem of two core incrementing the same variable at the
same time. Incrementing a variables is actually done in 3 steps: reading the
value from the memory, adding one, storing back the results. Since there is by
default non synchronization, the following sequence of actions for two core
and is possible: Read1 / Read2 / Add1 / Add2 / Store1 / Store2. Then the two
cores performs the same modification resulting in incrementing the variable
only once. This is called a data race and leads to nondeterministic wrong
result. To cope with those kinds of synchronization problems, Cilk provide the
notion of \emph{reducer} which are variables local to each thread which are
gathered (here added) when a thread is finishing its job.

A more important problem is that the cost of a recursive call in non
negligible. Using \tt{Cilk+} recursive calls instead of \tt{C++} calls makes
it even worse. The solution we use is to switch back to the non recursive
version using a stack when the genus is close to the target genus. This leads
to the following \tt{Cilk+} code:
\begin{listing}{1}
void explore(const monoid &m) {
  unsigned long int nbr = 0;
  if (m.genus < MAX_GENUS - STACK_BOUND) {
    auto it = generator_iter<CHILDREN>(m);
    while (it.move_next()) {
      auto child = remove_generator(m, it.get_gen()).
      cilk_spawn explore(child);
      nbr++;
    }
    cilk_results[m.genus] += nbr;
  }
  else explore_stack(m, cilk_results.get_array());
}
\end{listing}
Note that in our version, we found that \tt{STACK\_BOUND} optimal value was
around 10-12 for genus in the range $45\dots 67$ so that \tt{explore\_stack}
is used more than $99\%$ of the time. The \tt{Cilk+} recursive function does
actually very little work but ensure that the work is balanced between the
different cores.

@@

\subsubsection*{Avoiding dynamic allocation} Compared to the \textsc{Son}
algorithm, dynamic allocation cost orders of magnitude more. Therefore, during
the derecursived stack algorithm, we only allocate (on the system stack rather
than on the heap), the stack of monoid. No further allocation are done.

\subsubsection*{Avoiding all unnecessary copy} We also used a trick to avoid
copying from and to the top of the stack. Indeed, then main loop perform the
following sequences of operations:
\begin{algorithmic}[1]
  \State $\tt{S} \gets \tt{stack.top()}$
  \State \tt{stack.pop()}
  \For{all children $\tt{S}^\tt{x}$ of \tt{S}}
  \State $\tt{S.push}(\tt{S}^\tt{x})$
  \EndFor
\end{algorithmic}
If we use a stack of monoid, we can construct $\tt{S}^\tt{x}$ directly into
the stack memory but we have to copy the top of the stack to \tt{S}. Since the
average number of children is the golden ratio, this single copy in far from
being negligible. The trick is to use a level of indirection, replacing the
stack by an array of monoid and an array of indexes pointing to the array of
monoid. We keep the invariant that the array of indexes is a permutation of
the array of monoid. Now instead of copying \tt{S} out of the stack, we keep
it on the stack, pushing the children in the second position by exchanging the
indexes. Here is the relevant part of the code:
\begin{listing}{1}
  monoid data[MAX_GENUS-1], *stack[MAX_GENUS], *current;
  monoid **stack_pointer = stack + 1;
  for (ind_t i=1; i<MAX_GENUS; i++) stack[i] = &(data[i-1]);
  [...]
  while (stack_pointer != stack) {
      --stack_pointer;
      current = *stack_pointer;
      [...] for each children {
        *stack_pointer = *(stack_pointer + 1);
        [...] construct the children in **stack_pointer
        [...] using the parent in *current
        stack_pointer++;
        [...]
      }
      *stack_pointer = current;
  }
\end{listing}

@@

Due to the way, the C++ does its pointer arithmetic, even if the index in the
array are less than $3G$ and therefore fits in $8$~bits, we use $64$~bits
indexes (namely \verb|uint_fast64_t|) to avoid conversion and sign extension
when computing addresses of indexed element. This single standard trick save
20\% of speed.

@@

\begin{figure}[b]
\begin{tikzpicture}[scale=1.3]
        \begin{semilogyaxis}[xlabel=\small genus,ylabel=time,legend pos=north west,legend cell align=left]
        \addplot [color=blue,mark=x] coordinates {
            (20,0.02)
            (21,0.04)
            (22,0.07)
            (23,0.12)
            (24,0.2)
            (25,0.35)
            (26,0.61)
            (27,1.04)
            (28,1.76)
            (29,2.96)
            (30,4.96)
            (31,8.26)
            (32,13.77)
            (33,22.95)
            (34,38.22)
            (35,1250.88)
        };
       \addplot [color=black,mark=x] coordinates {
			(20,0.02)	
            (21,0.04)
            (22,0.06)
            (23,0.09)
            (24,0.16)
            (25,0.26)
            (26,0.45)
            (27,0.72)
            (28,1.25)
            (29,2.1)
            (30,3.39)
            (31,5.85)
            (32,9.25)
            (33,16.34)
            (34,27.23)
            (35,45.39)
            (36,75.58)
            (37,125.59)
            (38,203.56)
            (39,346.02)
            (40,557.51)
        };
         \addplot [color=red,mark=x] coordinates {
			(20,0.02)	
            (21,0.02)
            (22,0.04)
            (23,0.06)
            (24,0.1)
            (25,0.18)
            (26,0.29)
            (27,0.48)
            (28,0.81)
            (29,1.34)
            (30,2.25)
            (31,3.73)
            (32,6.2)
            (33,10.35)
            (34,17.11)
            (35,28.35)
            (36,46.99)
            (37,77.89)
            (38,128.53)
            (39,212.1)
            (40,349.98)
        };
        \addplot [color=magenta,mark=x] coordinates {
			(20,0.01)	
            (21,0.01)
            (22,0.01)
            (23,0.01)
            (24,0.02)
            (25,0.03)
            (26,0.05)
            (27,0.08)
            (28,0.12)
            (29,0.21)
            (30,0.35)
            (31,0.57)
            (32,0.97)
            (33,1.65)
            (34,2.69)
            (35,4.18)
            (36,7.36)
            (37,11.66)
            (38,20.35)
            (39,32.55)
            (40,74.46)
        };
	%\legend style={anchor=north,at={(20,1000)}}  
        \legend{\tiny breadth-expl,\tiny depth-expl,\tiny depth-expl-$\delta$,\tiny depth-expl-$\delta$-mmx}
        \end{semilogyaxis}
    \end{tikzpicture}
    \label{F:Time}
    \caption{Comparison of time executions of different exploration algorithms of the tree $\mathcal{T}^{\leq g}$ on an Intel \tt{i5-3570K} CPU with $8GB$ of memory. 
The scale is logarithmic in time. 
The first version is based on a breadth-first search exploration and the peak at genus~$35$ is due to the consumption of all the memory and use of swap. 
The second version uses a depth first search exploration.
The third is based on the second one with use of decomposition numbers, it corresponds to the algorithm given in Section~\ref{S:Algo}. 
The last one is an optimisation of the third one with MMX.}
\end{figure}

>>>>

\documentclass[reqno,11pt]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{pgfplots}
\usepackage{xspace}
\theoremstyle{plain}
\newtheorem{prop}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{defi}[prop]{Definition}
\newtheorem{nota}[prop]{Notation}
\newtheorem{exam}[prop]{Example}
\newtheorem{lem}[prop]{Lemma}
\newtheorem{fact}[prop]{Fact}

@@

\newcommand{\gr}[1]{{\color{gray} #1}}
%\newcommand{\card}{\#}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\Cilk}{\texttt{Cilk}\xspace}
\newcommand{\CilkP}{\texttt{Cilk++}\xspace}
\newcommand{\CPP}{\texttt{C++}\xspace}
\newcommand{\kkp}{{k+1}}
\newcommand{\gen}[1]{\left<#1\right>}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\newcommand{\MMX}{\texttt{MMX}\xspace}
\newcommand{\SIMD}{\texttt{SIMD}\xspace}
\newcommand{\SSE}{\texttt{SSE}\xspace}
\newcommand{\SSEV}{\texttt{SSE4.1}\xspace}
\newcommand{\NN}{\mathbb{N}}
\renewcommand{\tt}[1]{\texttt{#1}}
\newcommand{\sgnode}[1]{{\bf \left<#1\right>}}
\renewcommand{\ie}{\emph{i.e}}
\DeclareMathOperator{\Irr}{Irr}
\DeclareMathOperator{\App}{App}
\DeclareMathOperator{\card}{card}
\newcommand{\Jean}[1]{{\color{blue}#1}}
\newcommand{\Florent}[1]{{\color{green}#1}}
\newcommand{\Delete}[1]{}
\newcommand{\REM}[1]{\marginpar{\small #1}}
\newcommand{\TODO}[1]{\marginpar{\small To do: #1}}

@@

M.~Bras-Amor\'os used a depth first search exploration of the tree of numerical semigroups $\mathcal{T}$ up to a given genus.
This tree was introduced by J.C.~Rosales and al. in \cite{Rosales} and it is the subject of the Section~\ref{S:Tree}.
Starting with all the numerical semigroups of genus $49$ she obtained the number of numerical semigroups of genus $50$ in $18$ days on a pentium D runing at $3$GHz. 
In the package~\tt{NumericalSgs} \cite{NumericalSgps} of \tt{GAP} \cite{GAP}, M. Delgado together with P.A.~Garcia-Sanchez and J. Morais used the same method of exploration.

\Jean{Here we describe a new algorithm for the exploration of the tree of numerical semigroup $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$. 
The cornerstone of our method is a new combinatorial representation of numerical semigroup that is well suited for exploration of the tree~$\mathcal{T}$ and allow code optimization essentially based on use of vectorial instructions and parallelization.}

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
\Jean{The decription of our new representation of numerical semigroup is done in the third section.}
\Delete{In Section~\ref{S:Tree} we describe a new representation of numerical semigrouos that is well suited to the construction of the tree.}
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimisation of the algorithm introduced in Section~\ref{S:Algo}.

@@

To get the greatest speed from modern processor, we used several optimization
tricks, which we will elaborate in the following section:

\REM{C'est un peu moche c'est gros points, non? Des tirets peut etre.}
\begin{itemize}
\item Vectorization (\MMX, \SSE instructions sets) and careful memory alignment. 
\item Shared memory multi-core computing using \CilkP for low level
  enumerating tree branching;
\item Partially derecursived algorithm using a stack;
\item Avoiding all dynamic allocation during the computation: everything is
  computed ``in place'';
\item Avoiding all unnecessary copy (the cost of the \textsc{Son} algorithm is
  roughly the same as copying) !
\item Aggressive loop unrolling: the main loop is unrolled by hand using some
  kind of Duff's device.
\item Carefull choice of data type (\verb|uint_fast8_t| for decomposition
  number, vs \verb|uint_fast64_t| for all indexes).
\end{itemize}

@@

Assume for example that we want to construct the tree~$\mathcal{T}^{\leq 100}$
of all numerical semigroup of genus smaller than $100$.  In this case, the
representation of numerical semigroup given in Section~\ref{S:DecNumber} uses
decomposition numbers of integers smaller than~$300$.  By Lemma~\ref{L:RepN},
such a decomposition number is smaller than $151$ and requires~$1$ byte of
memory.  Thus at each \textbf{for} step of Algorithm~\textsc{Son}, the CPU
actually works on $1$ byte. However current CPUs usually work on $8$ bytes and
even on $16$~bytes using vector extensions. \REM{Peut etre rajouter qqchose sur l'AVX; en fait on devrait pouvoir utiliser l'AVX2.} The first optimization uses this
point.

@@

\noindent In this version we can see that the cursors \texttt{src} and
\texttt{dst} move at the same time and that the modification of the value
pointed by \texttt{dst} only needs to access the values pointed by \tt{src}
and \tt{dst}.  We can therefore work in multiple entries at the same time
without collision.  Current CPUs allow this thanks to the \SIMD technologies as
\MMX, \SSE, etc.  The acronym \SIMD~\cite{SIMD} stands for Single Operation
Multiple Data.  We used \SSEV~\cite{SSE, IntelSSE} technology as it allows
for the largest speedup. This need to respect some constraints in the memory
organization of the data, namely ``memory alignment''. Recall that an address
is 16~bytes aligned if it is a multiple of 16. \SSE memory access are much
faster for aligned access.

The computation of the children is then performed as follows. First, the
parent's decomposition numbers are copied in the children's using the
following \CPP code
\begin{listing}{1}
void copy_blocks(dec_blocks &dst, dec_blocks const &src) {
  for (ind_t i=0; i<NBLOCKS; i++) dst[i] = src[i];
}
\end{listing}
Here \tt{dec\_blocks} is a type for arrays of 16~bytes blocks whose size
\tt{NBLOCKS} are just large enough to store the decomposition numbers (that is
$3G$ rounded up to a multiple of 16). The instruction \tt{dst[i] = src[i]}
actually copy a full 16~bytes block.

Then the core of the \textbf{while} loop in the preceding algorithm is
translated as a \texttt{for} loop as follows (recall that $x$ denotes the generator of
the father which is to be removed in the children):
\begin{listing}{1}
start = x >> 4;      // index of the block containing x
shift = x & 0xF;     // offset of x inside the block
...                  // some specific instructions to
...                  // handle the first incomplete block.
for (long int i=start+1; i<NBLOCKS; i++) {
  block = load_unaligned_epi8(src + ((i-start)<<4) - shift);
  dst[i] -= ((block != zero) & one);
}
\end{listing}
The instruction \tt{load\_unaligned\_epi8} (specific to \SSE technology) loads
 16 consecutive entries of the decomposition number
of the parent (called \tt{src} monoid) in the variable \tt{block}.
Those entries will be used to compute
the entries $16i, \dots 16i+15$ of the children monoid. Since the removed
generator~$x$ is not necessarily a multiple of $16$, the data are not aligned
in memory, hence the use of a specific instruction. The \tt{zero}
(resp. \tt{one}) constants are initialized as 16 bytes equal to $0$
(resp. $1$). The comparison \tt{(block != zero)} therfore returns a block
which contains $0$ in the bytes corresponding the $0$ entries of block and
$255$ in the non zero one. This results is then bitwise anded with one so that
the instruction actually performs a 16~bytes parallel version of
\begin{algorithmic}
\State $\tt{dst} \gets \tt{dst} -
\textbf{ if } \tt{block} \neq 0 \textbf{ then } 1 \textbf{ else } 0$
\end{algorithmic}
which is equivalent to Lines \tt{10} to \tt{14} of Algorithm~\textsc{Son}.
\medskip

@@

\subsection{Parallel tree exploration using \CilkP}

Our second optimization is to use parallelism on exploration of the tree.
Today, CPU of personal computer have several cores (2, 4 or more).  The given
version of our exploration algorithm uses a single core and so a fraction only
of the power of a CPU. The idea here is that different branches of the tree can
be explored in parallel by different cores of the computer. The tricky part is
to ensure that each cores are busy, giving a new branch when a core is done
with a former one. Fortunately there is a technology called
\CilkP~\cite{CilkIntel, CilkRefman}
which is particularly well suited for those kind of problems.

\Cilk is a general-purpose language designed for multithreaded parallel
computing. The \CPP incarnation is called \CilkP. The biggest
principle behind the design of the \Cilk language is that the programmer should
be responsible for \emph{exposing} the parallelism, identifying elements that
can safely be executed in parallel; it should then be left to the run-time
environment, particularly the scheduler, to decide during execution how to
actually divide the work between processors.

The crucial thing is that two keywords are all that are needed to start
using the parallel features of \CilkP:
\begin{itemize}
\item\tt{cilk\_spawn}: used on a procedure call, indicates that the call can
  safely operate in parallel with the remaining code if the current
  function. Note that the scheduler is not obligated to run this procedure in
  parallel; the keyword merely alerts the scheduler that it can do so.
\item\tt{cilk\_sync}: indicates that execution of the current procedure cannot
  proceed until all previously spawned procedures have completed and returned
  their results to the current frame.
\end{itemize}
As a consequence, to get a parallel version of the recursive
Algorithm~\ref{A:ExpRec}, you only need to modify it as
\begin{algorithmic}[1]
\Procedure{ExploreRec}{\tt{S}, \tt{G}}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State\textbf{CilkSpawn} \Call{ExploreRec}{$\tt{S}^\tt{x}, \tt{G}$}
			\EndIf
		\EndFor
	\EndIf
\EndProcedure
\end{algorithmic}
We just tells \CilkP that the subtrees rooted at various children can be
explored in parallel.

The thing are actually only a little bit more complicated. First we have to
gather the results of the exploration. If we simply write
\begin{algorithmic}[1]
\State \tt{result}[$g(\tt{S})$] $\gets \tt{result}[g(\tt{S})] + 1$
\end{algorithmic}
then we face the problem of two cores incrementing the same variable at the
same time.
 Incrementing a variable is actually done in 3 steps: reading the
value from the memory, adding one, storing back the result. 
Since there is by
default non synchronization, the following sequence of actions for two cores
 is possible: Read1 / Read2 / Add1 / Add2 / Store1 / Store2. Then the two
cores performs the same modification resulting in incrementing the variable
only once. This is called a data race and leads to nondeterministic wrong
result. To cope with those kinds of synchronization problems, \Cilk provide the
notion of \emph{reducer} which are variables local to each thread which are
gathered (here added) when a thread is finishing its job.

A more important problem is that the cost of a recursive call in non
negligible. Using \CilkP recursive calls instead of \CPP calls makes
it even worse. The solution we use is to switch back to the non recursive
version using a stack when the genus is close to the target genus. This leads
to the following \CilkP code:
\begin{listing}{1}
void explore(const monoid &m) {
  unsigned long int nbr = 0;
  if (m.genus < MAX_GENUS - STACK_BOUND) {
    auto it = generator_iter<CHILDREN>(m);
    while (it.move_next()) {
      auto child = remove_generator(m, it.get_gen()).
      cilk_spawn explore(child);
      nbr++;
    }
    cilk_results[m.genus] += nbr;
  }
  else explore_stack(m, cilk_results.get_array());
}
\end{listing}
Note that in our version, we found that \tt{STACK\_BOUND} optimal value was
around 10-12 for genus in the range $45\dots 67$ so that \tt{explore\_stack}
is used more than $99\%$ of the time. The \CilkP recursive function does
actually very little work but ensure that the work is balanced between the
different cores.

@@

\subsubsection*{Avoiding dynamic allocation} Compared to the \textsc{Son}
algorithm, dynamic allocation cost orders of magnitude more. Therefore, during
the derecursived stack algorithm, we only allocate (on the system stack rather
than on the heap) the stack of monoid. No further allocation are done.

\subsubsection*{Avoiding all unnecessary copy} We also used a trick to avoid
copying from and to the top of the stack. Indeed, then main loop perform the
following sequences of operations:
\begin{algorithmic}[1]
  \State $\tt{S} \gets \tt{stack.top()}$
  \State \tt{stack.pop()}
  \For{all children $\tt{S}^\tt{x}$ of \tt{S}}
  \State $\tt{S.push}(\tt{S}^\tt{x})$
  \EndFor
\end{algorithmic}
If we use a stack of monoid, we can construct $\tt{S}^\tt{x}$ directly into
the stack memory but we have to copy the top of the stack to \tt{S}. Since the
average number of children is the golden ratio, this single copy is far from
being negligible. The trick is to use a level of indirection, replacing the
stack by an array of monoid and an array of indexes pointing to the array of
monoid. We keep the invariant that the array of indexes is a permutation of
the array of monoid. Now instead of copying \tt{S} out of the stack, we keep
it on the stack, pushing the children in the second position by exchanging the
indexes. Here is the relevant part of the code:
\begin{listing}{1}
  monoid data[MAX_GENUS-1], *stack[MAX_GENUS], *current;
  monoid **stack_pointer = stack + 1;
  for (ind_t i=1; i<MAX_GENUS; i++) stack[i] = &(data[i-1]);
  [...]
  while (stack_pointer != stack) {
      --stack_pointer;
      current = *stack_pointer;
      [...] for each children {
        *stack_pointer = *(stack_pointer + 1);
        [...] construct the children in **stack_pointer
        [...] using the parent in *current
        stack_pointer++;
        [...]
      }
      *stack_pointer = current;
  }
\end{listing}

@@

Due to the way, the \CPP does its pointer arithmetic, even if the index in the
array are less than $3G$ and therefore fits in $8$~bits, we use $64$~bits
indexes (namely \verb|uint_fast64_t|) to avoid conversion and sign extension
when computing addresses of indexed element. This single standard trick save
20\% of speed.

@@

\begin{figure}[b]
\begin{tikzpicture}[scale=1.3]
        \begin{semilogyaxis}[xlabel=\small genus,ylabel=time,legend pos=north west,legend cell align=left]
        \addplot [color=blue,mark=x] coordinates {
            (20,0.02)
            (21,0.04)
            (22,0.07)
            (23,0.12)
            (24,0.2)
            (25,0.35)
            (26,0.61)
            (27,1.04)
            (28,1.76)
            (29,2.96)
            (30,4.96)
            (31,8.26)
            (32,13.77)
            (33,22.95)
            (34,38.22)
            (35,1250.88)
        };
       \addplot [color=black,mark=x] coordinates {
			(20,0.02)	
            (21,0.04)
            (22,0.06)
            (23,0.09)
            (24,0.16)
            (25,0.26)
            (26,0.45)
            (27,0.72)
            (28,1.25)
            (29,2.1)
            (30,3.39)
            (31,5.85)
            (32,9.25)
            (33,16.34)
            (34,27.23)
            (35,45.39)
            (36,75.58)
            (37,125.59)
            (38,203.56)
            (39,346.02)
            (40,557.51)
        };
         \addplot [color=red,mark=x] coordinates {
			(20,0.02)	
            (21,0.02)
            (22,0.04)
            (23,0.06)
            (24,0.1)
            (25,0.18)
            (26,0.29)
            (27,0.48)
            (28,0.81)
            (29,1.34)
            (30,2.25)
            (31,3.73)
            (32,6.2)
            (33,10.35)
            (34,17.11)
            (35,28.35)
            (36,46.99)
            (37,77.89)
            (38,128.53)
            (39,212.1)
            (40,349.98)
        };
        \addplot [color=magenta,mark=x] coordinates {
			(20,0.01)	
            (21,0.01)
            (22,0.01)
            (23,0.01)
            (24,0.02)
            (25,0.03)
            (26,0.05)
            (27,0.08)
            (28,0.12)
            (29,0.21)
            (30,0.35)
            (31,0.57)
            (32,0.97)
            (33,1.65)
            (34,2.69)
            (35,4.18)
            (36,7.36)
            (37,11.66)
            (38,20.35)
            (39,32.55)
            (40,74.46)
        };
	%\legend style={anchor=north,at={(20,1000)}}  
        \legend{\tiny breadth-expl,\tiny depth-expl,\tiny depth-expl-$\delta$,\tiny depth-expl-$\delta$-mmx}
        \end{semilogyaxis}
    \end{tikzpicture}
    \label{F:Time}
    \caption{Comparison of time executions of different exploration algorithms of the tree $\mathcal{T}^{\leq g}$ on an Intel \tt{i5-3570K} CPU with $8GB$ of memory. 
The scale is logarithmic in time. 
The first version is based on a breadth-first search exploration and the peak at genus~$35$ is due to the consumption of all the memory and use of swap. 
The second version uses a depth first search exploration.
The third is based on the second one with use of decomposition numbers, it corresponds to the algorithm given in Section~\ref{S:Algo}. 
The last one is an optimisation of the third one with \MMX.}
\end{figure}
